---
title: "Game Analysis"
author: "Benjamin Calderaio, Jr."
date: "`r Sys.Date()`"
output:
  word_document: default
  # always_allow_html: true
  pdf_document: default
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    selfcontained: true
---
## Libraries
```{r}
library(DBI)
library(RMariaDB)
library(tidyr)
library(ggplot2)
library(plotly)
library(htmlwidgets)
library(reshape2)
library(dplyr)
```

## Environment Variables
Get variables from .Renviron
```{r, echo=FALSE,include=FALSE}
readRenviron("/Users/benc/Projects/r_proj/ds_notebooks/.Renviron")

# Set plotly to render offline for static documents
# plotly::config(p = plot_ly(), displayModeBar = FALSE)  # Example config for no mode bar
```

## Functions


```{r, echo=FALSE,include=FALSE}
db_con <- function() {
  con <- dbConnect(
    RMariaDB::MariaDB(),
    dbname = Sys.getenv("MYSQL_DBNAME"),
    host = Sys.getenv("MYSQL_HOST"),
    user = Sys.getenv("MYSQL_USER"),
    password = Sys.getenv("MYSQL_PASSWORD")
  )
  return(con)
}
```

### get_query
Returns a query string.
```{r, echo=FALSE,include=FALSE}
get_query <- function(table_name) {
  dbname <- Sys.getenv("MYSQL_DBNAME")
  query <- sprintf("SELECT * FROM %s.%s",dbname,table_name)
  return(query)
}
```

### get_csv_data
Gets csv data and returns a data frame.
```{r}
#' Get csv Data
#'
#' This function returns data from a database table.
#'
#' @param con The database connection
#' @param query The database query used to return the table rows
#' @return The results in a data frame
#' @examples
#' data <- get_data(con,qry)
get_csv_data <- function(fpath) {
  tryCatch({
    result_df <- read.csv(fpath)
    return(result_df)
  }, error = function(e) {
    print(paste("An error occurred:", e))
  })
}
```

### get_data
Gets data from a database table and returns a data frame.
```{r}
#' Get Database Data
#'
#' This function returns data from a database table.
#'
#' @param con The database connection
#' @param query The database query used to return the table rows
#' @return The results in a data frame
#' @examples
#' data <- get_data(con,qry)
get_data <- function(con,query) {
  tryCatch({
    result_df <- dbGetQuery(con, query)
    return(result_df)
  }, error = function(e) {
    print(paste("An error occurred:", e))
  }, finally = {
    dbDisconnect(con)
  })
}
```

### get_detail_fav_covers_by_year
Transforms detail spread for favorites and returns a data frame.
```{r}
get_detail_fav_covers_by_year <- function(leagyear,det_df) {
  leagyear_value <- leagyear
  df <- det_df %>%
    filter(leagyear == leagyear_value, isfav == 1, iscover == 1) %>%
    group_by(leagid,spread) %>%
    summarise(iscover_sum = sum(iscover, na.rm = TRUE), .groups = "drop") %>%
    ungroup()
  df <- as.data.frame(df)
  return(df)
}
```

### get_detail_fav_covers_v_dog_covers_by_year
Transforms detail spread for favorites v underdogs and returns a data frame.
```{r}
get_detail_fav_covers_v_dog_covers_by_year <- function(leagyear,det_df) {
  # det1_df <- read.csv('vw_gameteamresultsdetail.csv')
  leagyear_value <- leagyear
  df1 <- det_df %>%
    filter(leagyear == leagyear_value, isfav == 1, ispush == 0) %>%
    group_by(leagid,spread) %>%
    summarise(
      favcover = sum(iscover == 1, na.rm = TRUE),
      dogcover = sum(iscover == 0, na.rm = TRUE), .groups = "drop"
      ) %>%
    ungroup() %>%
    arrange(spread,leagid)
  df1 <- as.data.frame(df1)
  return(df1)
}
```

### create_plotly_spread_bar_chart
Creates a bar chart for favorites v underdogs.
```{r}
create_plotly_spread_bar_chart <- function(data) {
  df_nfl <- data %>% filter(leagid == 1)
  if(!is.numeric(df_nfl$spread)) {
    df_nfl$spread <- as.numeric((as.character(df_nfl$spread)))
  }
  df_nfl <- df_nfl %>% arrange(spread)
  df_nfl$spread <- factor(df_nfl$spread, levels = sort(unique(df_nfl$spread)))
  if (is.factor(df_nfl$spread)) {
    df_nfl$spread <- as.numeric(as.character(df_nfl$spread))
  }
  
  new_df <- data.frame(
    spread = df_nfl$spread,
    favorite = df_nfl$favcover,
    underdog = df_nfl$dogcover
  )
  # Sort data by offense total yards in descending order
  data_sorted_off <- new_df[order(-new_df$spread), ]
  
  # Reshape data for plotly, mapping variable names to readable labels
  data_long <- melt(data_sorted_off, id.vars = "spread", variable.name = "type", value.name = "covers")
  data_long$type <- ifelse(data_long$type == "favorite", "favorite", "underdog")
  
  # Plotly bar chart
  plot <- plot_ly(data = data_long,
                  x = ~spread, 
                  y = ~covers, 
                  color = ~type, 
                  type = "bar",
                  colors = c("favorite" = "blue", "underdog" = "orange")) %>%
    layout(title = "Favorite vs Underdog Spread cover",
           xaxis = list(title = "Spread", tickangle = -90),
           yaxis = list(title = "Covers"),
           barmode = 'group')  # This sets bars to be side by side
  
  print(plot)

  # htmlwidgets::saveWidget(plot, "fav_dog_spread.html", selfcontained = TRUE)
}
```

### create_ggplot_team_bar_chart
Creates a bar chart using ggplot for total offensive and defensive yards.
```{r}
create_ggplot_team_bar_chart <- function(data) {
  df_nfl <- data
  new_df <- data.frame(
    teamname = as.character(df_nfl$teamname),  # Convert teamname to character
    offense = df_nfl$offtotyds,
    defense = df_nfl$deftotyds
  )
  df <- suppressWarnings(tidyr::gather(new_df, yards, total, offense:defense))  # Create long format

  # Sort by total yards descending for each team (optional, if you want sorting)
  # df$teamname <- factor(df$teamname, levels = unique(df$teamname[order(df$total, decreasing = TRUE)]))

  plot <- ggplot(df, aes(teamname, total, fill=yards))
  plot <- plot +
    geom_bar(stat = "identity", position = 'dodge') +
    labs(title = "Offense vs Defense Total Yards", x = "Team", y = "Total Yards", fill = "Type") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("offense" = "blue", "defense" = "orange"))

  print(plot)

}
```

### create_ggplot_team_box_chart
Creates team box charts using ggplot for total offensive and defensive yard distrubtion.
```{r}
create_ggplot_team_box_chart <- function(data) {
  data_sorted_off <- data
  # Boxplot for defense total yards
  box_plot_d <- ggplot(data_sorted_off, aes(x = teamname, y = defense)) +
    geom_boxplot(fill = "red") +
    labs(title = "Distribution of Defense Total Yards",x = "Team", y = "Total Yards") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(box_plot_d)
  
  box_plot_o <- ggplot(data_sorted_off, aes(x = teamname, y = offense)) +
    geom_boxplot() +
    labs(title = "Distribution of Offense Total Yards",x = "Team", y = "Total Yards") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(box_plot_o)
}
```


### create_plotly_team_bar_chart
Creates a bar chart using plotly for total offensive and defensive yards.
```{r}
create_plotly_team_bar_chart <- function(data) {
  df_nfl <- data
  new_df <- data.frame(
    teamname = df_nfl$teamname,
    offense = df_nfl$offtotyds,
    defense = df_nfl$deftotyds
  )
  # Sort data by offense total yards in descending order
  data_sorted_off <- new_df[order(-new_df$offense), ]
  
  # Reshape data for plotly, mapping variable names to readable labels
  data_long <- melt(data_sorted_off, id.vars = "teamname", variable.name = "type", value.name = "yards")
  data_long$type <- ifelse(data_long$type == "offense", "offense", "defense")
  
  # Plotly bar chart
  plot <- plot_ly(data = data_long,
                  x = ~teamname, 
                  y = ~yards, 
                  color = ~type, 
                  type = "bar",
                  colors = c("offense" = "blue", "defense" = "orange")) %>%
    layout(title = "Offense vs Defense Total Yards",
           xaxis = list(title = "Team", tickangle = -45),
           yaxis = list(title = "Total Yards"),
           barmode = 'group')  # This sets bars to be side by side
  print(plot)
  
  create_ggplot_team_box_chart(data_sorted_off)
  # htmlwidgets::saveWidget(plot, "off_def_totyds.html", selfcontained = TRUE)
}
```

## main
R Markdown main method.
```{r}
main <- function(){
  csv_data <- get_csv_data('teamseasontotals_2024.csv')
  create_ggplot_team_bar_chart(csv_data)
  create_plotly_team_bar_chart(csv_data)
  
  con <- db_con()
  qry <- get_query(Sys.getenv("TABLE_GAMEDETAILS"))
  db_data <- get_data(con,qry)

  df <- get_detail_fav_covers_v_dog_covers_by_year(2024,db_data)
  create_plotly_spread_bar_chart(df)
}
```

## Output
```{r}
main()
```

